---
title: "Differential expression in acute myeloid leukaemia (GSE9476): t-test vs limma"
author: "Eden Black"
output:
  github_document:
    html_preview: true
    df_print: kable
---

This analysis identifies genes differentially expressed between acute myeloid
leukaemia (AML) and healthy bone marrow / peripheral blood in the public
microarray dataset GSE9476, and compares two ways of calling them — a per-gene
t-test and the `limma` linear-model framework. Expression values are
log2-transformed.

## Data import

```{r message = FALSE}
library(tidyverse)

# Read in datasets
AML_data = read.csv('data/GSE9476_series_matrix.txt',
                    header = TRUE,
                    sep = '\t',
                    fill = TRUE,
                    comment.char = '!')

AML_metadata = read.csv('data/GSE9476_meta_info.txt',
                        header = TRUE,
                        sep = '\t',
                        fill = TRUE)

# Convert metadata sample IDs and disease status to factors for later analysis
AML_metadata$SampleID = as.factor(AML_metadata$SampleID)
AML_metadata$Disease_status = as.factor(AML_metadata$Disease_status)
```

## Dataset overview

Sample numbers per group are counted from the metadata (possible because
`Disease_status` is a factor), and each probeset is one row of the expression
table.

```{r}
disease_status_counts = AML_metadata %>%
  group_by(Disease_status) %>%
  summarise(disease_status_count = n())

print(disease_status_counts)
```

```{r}
probeset_number = nrow(AML_data)

paste('There are', probeset_number, 'probesets in the dataset')
```

## Age as a covariate

Before comparing expression, I checked whether the two groups are balanced on
age, since an imbalance would be a potential confounder.

```{r}
t.test(Age ~ Disease_status, data = AML_metadata)
```

The groups differ significantly in age (p < 0.05), so age is a confounder to keep
in mind when interpreting these results.

## Exploratory analysis

The expression table is checked for missing values, pivoted to long format, and
joined to disease status so that each measurement is labelled AML or Control.

```{r}
sum(is.na(AML_data))
```

```{r}
# Tidy gene expression data by pivoting it into long format
AML_data_long = pivot_longer(
  data = AML_data,
  cols = GSM239170:GSM240509, # Exclude probe ID
  names_to = 'SampleID', # Match col name in metadata for easier joining
  values_to = 'Expression'
)

# Convert to factors for later analysis
AML_data_long$ID_REF = as.factor(AML_data_long$ID_REF)
AML_data_long$SampleID = as.factor(AML_data_long$SampleID)

str(AML_data_long)
```

```{r}
# Trim metadata to only SampleID (for join) and disease status
metadata_for_join = AML_metadata %>%
  select(SampleID, Disease_status)

# Join disease status from metadata to tidy gene expression data
AML_data_with_status = AML_data_long %>%
  left_join(metadata_for_join, by = 'SampleID')

str(AML_data_with_status)
sum(is.na(AML_data_with_status))
```

A per-sample boxplot of expression, coloured by disease status, gives a quick
view of overall distribution and any obvious outlier samples.

```{r, fig.width = 10, fig.height = 5}
ggplot(data = AML_data_with_status, aes(SampleID, Expression)) +
  geom_boxplot(aes(color = Disease_status, alpha = 0.6)) +
  scale_color_manual(values = c('AML' = 'red', 'Control' = 'blue')) +
  labs(title = 'Gene expression in AML and healthy samples',
       y = 'log2 Gene Expression',
       x = 'Sample ID',
       color = 'Disease Status') +
  scale_x_discrete(guide = guide_axis(angle = 90)) # Align x axis labels
```

## Per-gene summary statistics

For each probeset, summary statistics (mean, SD, min, max) are computed within
each group, then joined and used to derive a log2 fold change (mean AML − mean
Control).

```{r}
# Summarise AML data
AML_data_summary_stats = AML_data_with_status %>%
  filter(Disease_status == 'AML') %>%
  group_by(ID_REF) %>%
  summarise(mean_AML = mean(Expression),
            sd_AML = sd(Expression),
            min_AML = min(Expression),
            max_AML = max(Expression))

head(AML_data_summary_stats)
```

```{r}
# Summarise control data
control_data_summary_stats = AML_data_with_status %>%
  filter(Disease_status == 'Control') %>%
  group_by(ID_REF) %>%
  summarise(mean_Control = mean(Expression),
            sd_Control = sd(Expression),
            min_Control = min(Expression),
            max_Control = max(Expression))

head(control_data_summary_stats)
```

```{r}
# Join tables to create full summary stats for each gene
full_data_summary_stats = AML_data_summary_stats %>%
  full_join(control_data_summary_stats, by = 'ID_REF')

# Add column for fold change (mean_AML - mean_control)
full_data_summary_stats = full_data_summary_stats %>%
  mutate(Fold_change = mean_AML - mean_Control)

str(full_data_summary_stats)
sum(is.na(full_data_summary_stats))
```

A scatterplot of mean AML vs mean Control expression shows the relationship is
highly linear — most genes sit on the y = x line, and the genes that stray from
it are the candidates of interest.

```{r}
# Create scatter plot
ggplot(full_data_summary_stats, aes(mean_Control, mean_AML)) +
  geom_point(alpha = 0.4) +
  labs(title = 'Mean expression of genes in AML vs control samples',
       x = 'Mean log2 gene expression in control samples',
       y = 'Mean log2 gene expression in AML samples')
```

```{r}
# T-test for difference in overall mean gene expression in AML vs control
t.test(full_data_summary_stats$mean_AML,
       full_data_summary_stats$mean_Control)
```

As expected, overall gene expression is not shifted up or down between groups —
the signal of interest is in individual genes, not the global mean.

## Differential expression: per-gene t-tests

A t-test of expression by disease status is run for every probeset, and the
t-score and p-value are joined onto the summary-statistics table.

```{r}
t_test_table = AML_data_with_status %>%
  group_by(ID_REF) %>%
  mutate(t_score = t.test(Expression ~ Disease_status)$statistic, # Extract t-score into column
         p_value = t.test(Expression ~ Disease_status)$p.value) # Extract p-value into column
```

```{r}
full_data_with_t_test = t_test_table %>%
  select(ID_REF, t_score, p_value) %>%
  group_by(ID_REF) %>%
  full_join(full_data_summary_stats, by = 'ID_REF') %>%
  unique() %>% # Remove duplicate rows
  ungroup # Remove grouping to avoid problems later

str(full_data_with_t_test)
sum(is.na(full_data_with_t_test))
```

Before calling DEGs, probesets are restricted to those reliably expressed and not
saturated. Requiring mean expression above 3 in both groups leaves:

```{r}
filter_a = nrow(full_data_with_t_test %>%
                  filter(mean_AML > 3 & mean_Control > 3))

paste('There are', filter_a, 'probesets remaining after filter a')
```

Requiring maximum expression below 14 in at least one group leaves:

```{r}
filter_b = nrow(full_data_with_t_test %>%
                  filter(max_AML < 14 | max_Control < 14))

paste('There are', filter_b, 'probesets remaining after filter b')
```

Applying both criteria together — the more stringent, combined filter carried
forward — leaves:

```{r}
filter_c = nrow(full_data_with_t_test %>%
                  filter((mean_AML > 3 & mean_Control > 3) &
                           (max_AML < 14 | max_Control < 14)))

paste('There are', filter_c, 'probesets remaining after filter c')
```

The combined expression filter is applied, then DEGs are called as probesets with
p < 0.05 and |log2 fold change| > 1.

```{r}
DEG_data_filter_c = full_data_with_t_test %>%
  filter((mean_AML > 3 & mean_Control > 3) &
           (max_AML < 14 | max_Control < 14))

dim(DEG_data_filter_c)
```

```{r}
DEG_data_filtered = DEG_data_filter_c %>%
  filter(p_value < 0.05 & abs(Fold_change) > 1)

str(DEG_data_filtered)
sum(is.na(DEG_data_filtered))
```

```{r}
paste('There are',
      nrow(DEG_data_filtered),
      'differentially expressed genes in the dataset')
```

Splitting the DEGs by the sign of the fold change gives the up- and
down-regulated counts.

```{r}
upregulated_DEGs = DEG_data_filtered %>%
  filter(Fold_change > 0)

# Number upregulated
paste('There are',
      nrow(upregulated_DEGs),
      'upregulated DEGs in the dataset')

downregulated_DEGs = DEG_data_filtered %>%
  filter(Fold_change < 0)

# Number downregulated
paste('There are',
      nrow(downregulated_DEGs),
      'downregulated DEGs in the dataset')
```

The ten most up- and down-regulated DEGs by fold change:

```{r}
# 10 most upregulated DEGs with p-value, log2FC, and mean values
top_10_upregulated_DEGs = DEG_data_filtered %>%
  slice_max(order_by = Fold_change, n = 10) %>%
  select(ID_REF, p_value, Fold_change, mean_Control, mean_AML)

top_10_upregulated_DEGs
```

```{r}
# 10 most downregulated DEGs with p-value, log2FC, and mean values
top_10_downregulated_DEGs = DEG_data_filtered %>%
  slice_min(order_by = Fold_change, n = 10) %>%
  select(ID_REF, p_value, Fold_change, mean_Control, mean_AML)

top_10_downregulated_DEGs
```

## Differential expression: limma

The same DEG definition is applied using `limma`. The expression matrix is
prepared (probeset IDs as row names), a design matrix is built from disease
status, and a linear model is fitted with the contrast AML − Control.

```{r}
AML_data_limma = AML_data
rownames(AML_data_limma) = AML_data_limma$ID_REF # Apply ID_REF to row names
AML_data_limma$ID_REF = NULL # Remove old ID_REF column
```

```{r message = FALSE}
library(limma)

# Create design matrix based on metadata
design = model.matrix(~ 0 + AML_metadata$Disease_status)
colnames(design) = levels(AML_metadata$Disease_status)
```

```{r}
fit = limma::lmFit(AML_data_limma, design)

contrast.matrix = limma::makeContrasts(AML_vs_Control =
                                  AML - Control, levels = design)

fit2 = limma::contrasts.fit(fit, contrast.matrix)
fit2 = limma::eBayes(fit2)

limma_results = limma::topTable(fit2,
                   coef = 'AML_vs_Control',
                   adjust.method = 'BH',
                   sort.by = 'P',
                   n = Inf)

str(limma_results)
sum(is.na(limma_results))
```

The limma results are joined to the summary-statistics table and put through the
same filters as the t-test approach, for a like-for-like comparison: mean > 3 in
both groups, max < 14 in at least one group, p < 0.05, and |log2 fold change| > 1.

```{r}
# Create ID_REF column in limma results to join on
limma_results$ID_REF = rownames(limma_results)

# Join limma_results with summary stat data for further filtering
full_data_with_limma = limma_results %>%
  select(ID_REF, logFC, P.Value, adj.P.Val) %>%
  group_by(ID_REF) %>%
  full_join(full_data_summary_stats, by = 'ID_REF') %>%
  unique() %>% # Remove duplicate rows
  ungroup # Remove grouping to avoid problems later

str(full_data_with_limma)
sum(is.na(full_data_with_limma))
```

```{r}
limma_results_filtered = full_data_with_limma %>%
  filter((mean_AML > 3 & mean_Control > 3) &
           (max_AML < 14 | max_Control < 14) &
              P.Value < 0.05 & abs(logFC) > 1)

str(limma_results_filtered)
sum(is.na(limma_results_filtered))
```

```{r}
paste('There are', nrow(limma_results_filtered), 'DEGs in the dataset')
```

## Comparing the two methods

An inner join between the two DEG sets counts the genes identified by both
methods.

```{r}
# Identify DEGs in both methods using inner join
DEGs_limma_and_t_test = DEG_data_filtered %>%
  inner_join(limma_results_filtered, by = 'ID_REF')

overlapped_gene_count = nrow(DEGs_limma_and_t_test)

paste('There are', overlapped_gene_count, 'overlapped genes')
```

The overlap is visualised as a Venn diagram.

```{r}
library(VennDiagram)

# Create Venn diagram
grid.newpage()
grid.draw(venn.diagram(x = list(t_test = DEG_data_filtered$ID_REF,
                                limmma = limma_results_filtered$ID_REF),
                        category.names = c('T-test', 'Limma'),
                        fill = c(alpha('red',0.3),
                                  alpha('blue',0.3)),
                       filename = NULL))
```

The two methods agree on the large majority of DEGs, each identifying a small
number the other does not. Those method-exclusive genes can be retrieved with
anti-joins.

```{r}
t_test_exclusive_DEGs = DEG_data_filtered %>%
  anti_join(limma_results_filtered, by = 'ID_REF')

print(t_test_exclusive_DEGs)
```

```{r}
limma_exclusive_DEGs = limma_results_filtered %>%
  anti_join(DEG_data_filtered, by = 'ID_REF')

print(limma_exclusive_DEGs)
```

## Session information

```{r}
sessionInfo()
```
